HMM

### 2015-12-29 paper3.Rnw
more on methods

\begin{document}
%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
\maketitle

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}.As \citeauthor{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e.  the
assumption of temporal homogeneity in behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission (FWC) staff using
trained hounds and houndsmen. Of these panthers, 13 had 
sufficient data to be used by \cite{kerk2015hidden}. Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is to larger data sets \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.

We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the 'true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is more convenient to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998},
but we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

\todo{This description of the V. algorithm is not coherent; need to give
just a \emph{brief} definition here (since we are going to use it) and criticize/
explain shortcomings later}

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$, the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$ and the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ at state $\state_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$, the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\state_{k}$ at time $t=k$.

\item[Emission probability] $P({\boldmath y_{k}} | \state_{k},\cov_{k})$, a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting.

We modeled three types of transition complexity instead of time homogeneous transition; extension of ``Switch with covariates" by \cite{morales_extracting_2004}. For $n$ possible movement states, we have a $n \times n$ matrix that defines the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$ at time $t$ as the following multinomial logistic:
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

where $\eta_{ij}(t)$ can be modelled as the following: 
\begin{enumerate}
\item Multiple block transition:  The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm}b_{m,t}
$$
where $a_{ijm}$ are parameters, and $b_{m,t}$ is a binary indicator (1 to the time block at the corresponding time $t$, and 0 otherwise).  
\item Quadratic functional transition with a time covariate:
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
$$
\item Sinusoidal functional transition with a time covariate:
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos(\frac{2\pi t}{24})+b_{ij3}\sin(\frac{2\pi t}{24})
$$
\end{enumerate}

We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches. First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\Delta$ BIC = BIC - min(BIC)). Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity. Lastly, autocorrelation comparison between the observed and simulated data.

The package depmixS4 \citep{visser2010depmixs4} for R
(R Development Core Team 2011) was used for fitting covariate dependent transition HMM
models, simulating states and step lengths using the
estimated parameters, and for estimating states with the Viterbi
algorithm.

@@

As a compliment, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.

>>>>

\begin{document}
\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}

%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
\maketitle

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states is
approximately constant. Changes in behavioural mode switching 
\todo{do we want to call these transition probabilities (throughout)?}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in switching behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of behavioural modes.

Our second goal, therefore, is to explore whether
\citeauthor{kerk2015hidden}'s results might be driven at
least in part by structural problems with the HMM, i.e. the
assumption of temporally homogeneous behaviour.  When large data sets
are available, information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of behavioural modes), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of behavioural modes; we also
fit models to simulated data with varying numbers of behavioural modes
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets. \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats?}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work}

@@

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent behavioural modes.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever}
In general, BIC values decreased as the number of states
increased from 
three to six states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent}
suggesting that the six-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

\todo{The following description of the V. algorithm is not coherent; need to give
just a \emph{brief} definition here (since we are going to use it) and criticize/
explain shortcomings later}

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $\mathbf Y = \obs_{1},..., \obs_{T}$ and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model parameters {$\boldsymbol \theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \times \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$; the vector of observation densities $\obs_{1}$ conditions on covariates $\cov_{1}$ and state $\state_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observation density $\obs_{k}$ condition on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
While naive approaches to this averaging process there are several
more efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \citep{zucchini_hidden_2009};
we use the implementations in the {\tt depmixS4} package for R
\citep{visser2010depmixs4,R}.

We considered 3 (biologically interpretable number of states identified: (1) resting mode, characterized by short step length; (2) Moderately active mode characterized by intermediate step lengths, and (3) Traveling mode, characterized by long step lengths) to 6 (maximum number of states by BIC selection from time homogeneous transition HSMM) underlying movement states in \citet{kerk2015hidden} for HMM fitting. \todo{where do we actually focus on these three states?  Think this could probably be introduced in the Discussion, if we need it}

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  However, this matrix needs to respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1. Therefore, as is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is unnecessary (i.e. we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

We considered three different models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004}. \todo{second clause isn't clear; expand?}.

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta or  indicator variable (1 for the time block at the corresponding time $t$, and 0 otherwise).  
\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour, with an additional diurnal continuity constraint that $\eta_{ij}(0)=\eta_{ij}(24)$ \todo{am I right/do we actually impose this constraint? give a bit more detail ...}
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
$$
\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, \todo{apparently true from eqs here, but not true if we constrain the quadratic model??} but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$
\end{description}

\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.}

\subsection{Model evaluation}

\todo{Give more details about parameters, structure of simulations, etc.}

The {\tt depmixS4} package
was used for fitting covariate dependent transition HMM
models, simulating states and step lengths using the
estimated parameters, and for estimating states with the Viterbi
algorithm.

\todo{describe the Viterbi algorithm here, and the autocorrelation function
computation - everything that appears in the results should be described
explicitly here}

We assessed the fit of the time homogeneous and inhomogeneous transition HMMs to 4 of the 13 Florida panthers with the most data using three approaches. First, goodness of fit of different number of states by scaled BIC within each type of transition complexity ($\dbic$ = BIC - min(BIC)). Secondly, the average distance by time of day comparison of observed and simulated data from each type of transition complexity. Lastly, autocorrelation comparison between the observed and simulated data.

\todo{give at least a hint of why/what each of these three 
approaches ($\dbic$, hourly averages, ACF) will show}

@@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit.  \todo{if FMMs are here they should be in methods} Looking at figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.
